---
title: "homework-03"
author: "Carlo Lapayese-Calderon"
date: 2026-06-04
format: pdf
editor: visual
---

# Setup

```{r}

suppressPackageStartupMessages(library(tidyverse))

suppressPackageStartupMessages(library(here))

suppressPackageStartupMessages(library(readxl))

suppressPackageStartupMessages(library(janitor))

kelp <- read_csv(here("data", "temp-kelp.csv"))

bird <- read_csv(here("data", "bird_weather_observations.csv"))

```

# Problem 1. Giant kelp fronds

## 1a

To evaluate the relationship between ocean temperature and giant kelp frond elongation rate, I would use both Pearson correlation and simple linear regression. Pearson correlation measures the strength and direction of a linear relationship between two continuous variables, while simple linear regression models how kelp growth rate changes as a function of temperature and provides an interpretable slope.

## 1b

```{r}

# ggplot will give us our default plot using kelp
# object and then aes will help us specify the
# x-axis and y-axis of interest.
ggplot(kelp, aes(x = temp_c, y = kelp_elong)) +
  # geom_point will create the actual
  # scatter plot, and we can use steelblue as the
  # color and alpha to reduce opacity.
  geom_point(color = "steelblue", alpha = 0.7) +
  # This lets us rename our x-axis and y-axis to
  # something much cleaner.
  labs(
    x = "Ocean temperature (°C)",
    y = "Kelp frond elongation rate (cm/day)",
    # title allows us to name title.
    title = "Relationship between temperature and kelp growth"
  ) +
  # Nothing more than a theme that I found to be
  # pleasing to the eyes.
  theme_minimal()

```

## 1c

### Check assumptions

```{r}

# Fit simple linear regression model.
kelp_lm <- lm(kelp_elong ~ temp_c, data = kelp)

# Diagnostic plots for assumptions
par(mfrow = c(2, 2))
plot(kelp_lm)

```

The assumptions of simple linear regression that were checked include linearity, normality of residuals, and homoscedasticity. These were assessed using diagnostic plots from the linear model, including residuals vs fitted values and Q-Q plots. The scatterplot from Part 1b also suggested an approximately linear, negative relationship between temperature and kelp growth, although there was noticeable scatter around the trend.

### Run statistical tests

```{r}

# Pearson correlation test
cor.test(kelp$temp_c, kelp$kelp_elong)

# View model summary.
summary(kelp_lm)

```

## 1d

## 1e

## 1f

# Problem 2. Personal data

## 2a

## 2b

# Problem 3. Affective visualization

## 3a

## 3b

## 3c

## 3d

## 3e

# Problem 4. Statistical critique

## 4a

## 4b

## 4c

## 4d
