---
title: "homework-03"
author: "Carlo Lapayese-Calderon"
date: 2026-06-04
format: pdf
editor: visual
---

# Setup

```{r}

suppressPackageStartupMessages(library(tidyverse))

suppressPackageStartupMessages(library(here))

suppressPackageStartupMessages(library(readxl))

suppressPackageStartupMessages(library(janitor))

kelp <- read_csv(here("data", "temp-kelp.csv"))

bird <- read_csv(here("data", "bird_weather_observations.csv"))

```

# Problem 1. Giant kelp fronds

## a

To evaluate the relationship between ocean temperature and giant kelp frond elongation rate, I would use both Pearson correlation and simple linear regression. Pearson correlation measures the strength and direction of a linear relationship between two continuous variables, while simple linear regression models how kelp growth rate changes as a function of temperature and provides an interpretable slope.

## b

```{r}

```


## c

## d

## e

## f

# Problem 2. Personal data

## a

## b

# Problem 3. Affective visualization

## a

## b

## c

## d

## e

# Problem 4. Statistical critique

## b

## c

## d
