---
title: "homework-03"
author: "Carlo Lapayese-Calderon"
date: 2026-06-04
format: pdf
editor: visual
---

# Setup

```{r}

suppressPackageStartupMessages(library(tidyverse))

suppressPackageStartupMessages(library(here))

suppressPackageStartupMessages(library(readxl))

suppressPackageStartupMessages(library(janitor))

kelp <- read_csv(here("data", "temp-kelp.csv"))

bird <- read_csv(here("data", "bird_weather_observations.csv"))

```

# Problem 1. Giant kelp fronds

## a

To evaluate the relationship between ocean temperature and giant kelp frond elongation rate, I would use both Pearson correlation and simple linear regression. Pearson correlation measures the strength and direction of a linear relationship between two continuous variables, while simple linear regression models how kelp growth rate changes as a function of temperature and provides an interpretable slope.

## b

```{r}

# ggplot will give us our default plot using kelp
# object and then aes will help us specify the
# x-axis and y-axis of interest.
ggplot(kelp, aes(x = temp_c, y = kelp_elong)) +
  # geom_point will create the actual
  # scatter plot, and we can use steelblue as the
  # color and alpha to reduce opacity.
  geom_point(color = "steelblue", alpha = 0.7) +
  # This lets us rename our x-axis and y-axis to
  # something much cleaner.
  labs(
    x = "Ocean temperature (°C)",
    y = "Kelp frond elongation rate (cm/day)",
    # title allows us to name title.
    title = "Relationship between temperature and kelp growth"
  ) +
  # Nothing more than a theme that I found to be
  # pleasing to the eyes.
  theme_minimal()

```


## c

## d

## e

## f

# Problem 2. Personal data

## a

## b

# Problem 3. Affective visualization

## a

## b

## c

## d

## e

# Problem 4. Statistical critique

## b

## c

## d
